MS Technical Paper: \\ Placement Algorithms for Heterogenous FPGAs

### 2025-03-24 latex/paper/paper.tex
work on paper

\section{Design Flow}
A typical FPGA design flow goes like the following. First, the developer writes the design entry in a high level Hardware Description Language (HDL) like Verilog or VHDL.
\begin{itemize}
\item Synthesis (Vivado default)
\item Placement (the focus of this paper)
    \begin{itemize}
        \item \textbf{Prepacking}:\\
        Traverse the raw EDIFNetlist from synthesis and identify patterns of cells that should be grouped together.
        This can be CARRY chains, DSP cascades, LUTFF pairs, etc.
    \item \textbf{Packing}:\\
        Take the prepacked cell groups and pack them into SiteInsts.
        Intra\-Site routing.
    \item \textbf{Placement}:\\
        Take the SiteInsts and optimally place them onto physical Sites.
    \end{itemize}
\item Routing
    Inter\-Site routing
\end{itemize}

>>>>

\section{Introduction}

    Field-Programmable Gate Arrays (FPGAs) have witnessed unprecedented growth in capacity and versatility, driving significant advances in computer-aided design (CAD) and electronic design automation (EDA) methodologies. 
    Since the early-to-mid 2000s, the stagnation of single-processor performance relative to the rapid increase in integrated circuit sizes has led to a design productivity gap, where the computational effort for designing complex chips continues to rise. 
    In FPGA CAD flows—which traditionally encompass logic synthesis, placement, routing, and bitstream generation—the placement stage has emerged as one of the most time-consuming processes. 
    Inefficiencies in placement not only extend design times from hours to days, thereby elevating cost and reducing engineering productivity, but also limit the broader adoption of FPGAs by software engineers who expect compile times akin to those of conventional software compilers.

    In this paper, we recreate and experiment with established placement algorithms by leveraging the Xilinx's RapidWright, which is a semi-open-source API that offers backend access to Xilinx's industry-standard FPGA environment, Vivado. We implement a simulated annealing placer for Xilinx's 7-series FPGAs, with an emphasis on minimizing total wirelength while mitigating runtime. Our implementation is organized into three consecutive substages. The \textbf{prepacking} stage involves traversing a raw EDIF netlist to identify recurring cell patterns—such as CARRY chains, DSP cascades, and LUT-FF pairs—that are critical for efficient mapping and legalization. In the subsequent \textbf{packing} stage, these identified patterns, along with any remaining loose cells, are consolidated into SiteInst objects that encapsulate the FPGA’s discrete resource constraints and architectural nuances. Finally, the \textbf{placement} stage employs a simulated annealing (SA) algorithm to optimally assign SiteInst objects to physical sites, aiming to minimize total wirelength while adhering to the constraints of the 7-series architecture.

    Simulated annealing iteratively swaps placement objects, in our case, SiteInsts, guided by a cost function that decides which swaps should be accepted or rejected. Hill climbing is permitted by occasionally accepting moves that increase cost, in hope that such swaps may later lead to a better final solution. SA remains a popular approach in FPGA placement research due to its simplicity and robustness in handling the discrete architectural constraints of FPGA devices. While SA yields surprisingly good results given relatively simple rules, it is ultimately a heuristic and stochastic approach that explores the vast placement space by making random moves. Most of these moves will be rejected, meaning that SA must run many iterations, usually hundreds to thousands, to arrive at a desirable solution.

    In the ASIC domain, where placers must handle designs with millions of cells, the SA approach has largely been abandoned in favor of analytical techniques, owing to SA's runtime and poor scalability. Modern FPGA placers have also followed suit, as new legalization strategies allow FPGA placers to leverage traditionally ASIC placement algorithms and adapt them to the discrete constraints of FPGA architectures. While this paper does not present a working analytical placer, it will explore ways to build upon our existing infrastructure (prepacker and packer) to replace SA with AP.

\section{Xilinx 7-Series Architecture}
    Before any work can be done on a placer, one must consider what is being placed and upon what it is being placed. \\
    (Square peg square hole analogy) \\
    (Show the arch hierarchy). \\

    how to organize this? \\

    Device hierarchical objects: \\
        Super Logic Region, Clock Region, Tile, Site, BEL \\

    Device connection objects: \\
        Switchbox, PIP, BELPin, SitePin \\

    Device object instances: \\
        SitePinInst, SiteInst, CellPin

    EDIF objects: \\
        EDIFNet, EDIFCell, EDIFPort \\
        EDIFHierNet, EDIFHierCell, EDIFHierPort \\
        EDIFHierCellInst, EDIFHierPortInst \\

\section{RapidWright API}
    RapidWright represents the architecture constructs as faithfully as possible

    The RapidWright API offers a user-friendly toolkit to take pre-implemented designs, that is, designs that have already been placed and/or routed by Vivado, and make custom optimizations to fit complex design criteria. 
    RapidWright has several "BlockPlacers", a general Router, but no general Placer. 
    We use this API to implement a crude general placer, in our case, a simulated annealer. 
    (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain).

\section{Simulated Annealing}
    (Describe the algorithm, then in context of 7-Series arhictecture). \\

\section{Analytical Placement}
    Talk about HeAP.
